Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations

### 2015-05-22 Paper.tex
Updated images increasing picture quality and making sure that every was obtained with the latest version of ANIMO. Updated minor details in the text.

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}
\pgfdeclareimage[width=\textwidth]{100TNF0EGF}{images/TNFa-EGF/100TNFa-0EGF/100TNF_0EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=\textwidth]{0TNF100EGF}{images/TNFa-EGF/0TNFa-100EGF/0TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=\textwidth]{100TNF100EGF}{images/TNFa-EGF/100TNFa-100EGF/100TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=\textwidth]{CircadianComparison}{images/Circadian/ANIMOvsODE-crop}

@@

The output of the ANIMO model in Figure~\ref{fig:drosophila-model} matches the original ODE model.
In particular, the oscillations in both models show the same periods and phases (see Suppl. Fig~\ref{suppl-fig:grafici-drosophila}).
% Due to the compositional nature of \tas\, ANIMO allows for intuitive \emph{in silico} knock-out experiments,
% by right-clicking a node in the model and disabling it. Such experiments have been done
% before~\cite{drosophila-ode-model} and give similar results in our model.
Details of the comparison
between the ANIMO model and the original ODE model is given in Supplementary Section~\ref{suppl-sec:animo-drosophila}.
% \\
% {\Large [[Insert description of K.O. experiments with Drosophila model]]}

@@

\def\largeModelScale{0.18}%0.15}%0.155}%0.27}
\def\legendScalaColori{0.21}%0.18}%0.21}
\def\legendScalaForme{0.21}%0.18}%0.21}
\def\scalaGrafici{0.0709}%0.13}
\begin{figure*}[!htpb]
\centering
%  \subfloat[\label{fig:large-model-no-hypotheses}]{\includegraphics[scale=\largeModelScale]{images/large_network_merged_no_hypotheses}}\\
% %\subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme}}\\ \subfloat{\includegraphics[scale=\legendScalaForme]{images/legenda_colori}}\\
% \subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme_e_colori}}\\
% \addtocounter{subfigure}{-1}
% \subfloat[\label{fig:large-model-graph1}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_TNF0_ho_hyp_IL-1a3}}
% \subfloat[\label{fig:large-model-graph3}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_no_hyp_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_merged_no_hypotheses_oneImage}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-no-hypotheses}}
The model for the merged TNF$\alpha$ and EGF pathways. Node colours represent the
activity level of the corresponding modelled reactants at time $t = 15$ minutes after
a stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph1}}~
Modelled production of IL-1$\alpha$ after stimulation with 100 ng/ml TGF$\alpha$ (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph3}}~
Modelled activation of JNK1 and MK2 after stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-all}
\end{figure*}

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_oneImage}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

There are other nodes for which the experimental data deviates from the model in one or more of the experimental conditions.
A comparison between model and experimental data can be found in Figures~\ref{fig:differences1}, \ref{fig:differences2} and~\ref{fig:differences3}.
Comparing these results with the ones from~\cite{pathway-fuzzy} shows a better fit of the ANIMO model, which is also intrinsically
more precise, being more mechanistic in nature (see Fig.~\ref{fig:animo-spectrum}).
A complete deciphering of the signalling events
in this biological system is outside the scope of this paper. Instead, we illustrated how interactive modelling of the dynamic behaviour
of a signal transduction network can be used to extend previous pathway topologies and can lead to the generation of novel hypotheses.

@@

Timed Automata have been shown to be a powerful formalism to model biological processes
~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together
these locations represent the current state of the biological system. Each timed automaton can have one or more local clocks
associated to it, allowing temporal control of transitions between locations. These transitions are used to
represent interactions between molecules. Fast interactions take less time than slow interactions
to perform an activation or inhibition step. We have previously described in detail how approximated
scenarios~\cite{animo-ieee} (see Section~\ref{subsec:abstractions}) can be used to calculate the timing of molecular
interactions to give a description of network dynamics. Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas.
This model describes the activation of ERK by MEK.

@@

\begin{figure*}[!htbp]
\centering
%  \\
% \subfloat{\includegraphics[scale=\legendGraphScale]{images/small-model-1g_legenda}}\addtocounter{subfigure}{-1}\subfloat[\label{fig:small-model-first}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model1f}}
% %\subfloat[\label{fig:small-model-second}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model2f}}
% \subfloat[\label{fig:small-model-third}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model3f}}
% \subfloat[\label{fig:small-model-fourth}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model4g}}
%    \\
% \subfloat[\label{fig:small-model-first-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph1m_riga}}
% %\subfloat[\label{fig:small-model-second-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph2m_riga}}
% \subfloat[\label{fig:small-model-third-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph3n_riga}}
% \subfloat[\label{fig:small-model-fourth-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph4o_riga}}
  \includegraphics[width=0.9\textwidth]{images/small_model_oneImage}
  \caption{
%{\bf (\protect\subref*{subfig:mek-erk} - \protect\subref*{subfig:mek})}
%\\
%{\bf (\protect\subref*{fig:small-model-first} - \protect\subref*{fig:small-model-fourth-graph})}
\csentence{Construction of an ANIMO model of signal transduction
events in human colon carcinoma cells upon stimulation with 100 ng/ml TNF$\alpha$.}
Graphs below show the dynamic behaviour of the corresponding models above, comparing it to the measured
activity values from~\cite{pathway-compendium} (error bars represent the standard deviation).
On the vertical axis, ``100'' represents the maximum protein activity in the complete experiment.
A red vertical line in each graph highlights an arbitrary time point in the time course:
nodes in the corresponding model are coloured according to their activity at that time point.
{\bf(A,~D)} %{\bf (\protect\subref*{fig:small-model-first}, \protect\subref*{fig:small-model-first-graph})}~
Basic model showing direct activation of JNK1 and MK2 by TNF$\alpha$.
No peak dynamics are observed because no inactivating processes are present.
%{\bf (\protect\subref*{fig:small-model-second}, \protect\subref*{fig:small-model-second-graph})}~The model with added phosphatases.
{\bf(B,~E)} %{\bf (\protect\subref*{fig:small-model-third}, \protect\subref*{fig:small-model-third-graph})}~
The model after addition of inactivating phosphatases and a negative feedback loop that down-regulates TNFR. Note that adding TNFR internalization or phosphatases alone would not be enough to reproduce activity peaks.
{\bf(C,~F)} %{\bf (\protect\subref*{fig:small-model-fourth}, \protect\subref*{fig:small-model-fourth-graph})}~
The model after addition of IKK, IL1-secretion (abstracting
the autocrine IL-1 signalling described in~\cite{pathway-autocrine}), Casp8 and Casp3, showing the late response to TNF$\alpha$ signalling.
As the data set did not contain values for cleaved caspase-3, but only for its non-cleaved precursor pro-caspase-3,
we computed the {\sf Casp3\_{}data} series as $100\% - [\mbox{\sf pro-Casp3}]$.\label{fig:small-model}}
\end{figure*}
%\end{floatingfigure}

@@

\begin{center}
\begin{longtable}{lll}
\caption{Explanation of the abbreviated names referring to molecular species
in the main text.}\label{suppl-tab:names} \\
%\begin{tabular}{lll}
\hline
\noalign{\vskip 2mm} {\bfseries Abbreviation} & {\bfseries Full name} \\[2mm]
\hline
\endfirsthead

\multicolumn{3}{c}%
{\scriptsize{\bfseries \tablename\ \thetable{}} \--\ continued from previous page} \\
\hline
\noalign{\vskip 2mm} {\bfseries Abbreviation} & {\bfseries Full name} \\[2mm]
\hline
\endhead

@@

Detailed biochemical reaction mechanisms are abstracted to \emph{interactions}, which can
represent either activations ($\rightarrow$) or inhibitions ($\dashv$\hspace{0.1em}).
This aggregation of elementary reactions into single interaction steps reduces the number of kinetic
parameters involved, while preserving cause-and-effect relationships.
For example, consider a reaction in which enzyme $E$ phosphorylates and activates substrate $S$,
transferring a phosphate group from a molecule of ATP to a molecule of $S$. Biochemically, this reaction
can be represented as
$$
\mbox{\it E} + \mbox{\it S} + \mbox{\it ATP} \rightleftarrows \mbox{\it ES} + \mbox{\it ATP} \rightarrow \mbox{\it ES}^{\mbox{\scriptsize \it P}} + \mbox{\it ADP} \rightleftarrows \mbox{\it E} + \mbox{\it S}^{\mbox{\scriptsize\it P}} + \mbox{\it ADP},
$$
with conservation condition $\mbox{\it S} + \mbox{\it S}^{\mbox{\scriptsize\it P}} = \mbox{constant}$ and $\mbox{\it ATP} + \mbox{\it ADP} = \mbox{constant}$.\\
Under the assumption of ATP constantly being replenished by the cell, this reaction is abstracted in ANIMO to the corresponding interaction
$$
\mbox{\it E} \rightarrow \mbox{\it S}.
$$
Each occurrence of the interaction $E \rightarrow S$ will increase the activity level of $S$ by one discrete step.
Since the activity level is defined as the active fraction of a molecular species, an increase in the active fraction
implies a decrease in the inactive fraction. Hence, the original conservation condition is automatically
satisfied.
The interaction rate, $R$, depends on the activity levels of the reactants involved and on a single kinetic
parameter $k$ that is set by the user.
The three available interaction scenarios can be interpreted as abstracted kinetic rate laws:
\begin{enumerate}
  \item $R = k \times [E]$: the interaction rate depends only on the activity level of the upstream node.
  \item $R = k \times [E] \times [1 - S]$ (activations) or $R = k \times [E] \times [S]$ (inhibitions): the rate
  depends on the activity levels of both the upstream and downstream participants. Activations depend on the
  presence of inactive substrate, \emph{[1 - S]}, whereas inhibitions depend on the level of active substrate,
  \emph{[S]}.
  \item $R = k \times [E_1] \times [E_2]$: this scenario can be used when the activation or inhibition
  of a downstream node depends on the simultaneous activity of two upstream nodes. This scenario is comparable to an
  \emph{AND-gate} in Boolean logic.
\end{enumerate}
We will show in Section~\ref{sec:results} that the abstraction proposed here preserves ample
expressivity to capture the dynamic behaviour of a biological network.

\subsection{ANIMO}
The modelling approach described in Section~\ref{subsec:abstractions} and Section~\ref{subsec:timed-automata}
is implemented in the
software tool ANIMO (Analysis of Networks with Interactive MOdelling, \cite{animo-ieee}) as
a plug-in to the network visualization tool Cytoscape~\cite{cytoscape}. The visual interface of Cytoscape
makes the construction, expansion and rewiring of a network topology a fast and user-friendly process.

@@

ANIMO automatically translates the user input to a \tas\ model, which is then simulated with the model
checking tool UPPAAL~\cite{uppaal}. The results are subsequently parsed and translated to a graph that shows
the dynamic behaviour of nodes in the network.
A schematic overview of this process is given in Supplementary Section~\ref{suppl-sec:animo-ta}.
No training or prior knowledge on the use of \tas\ or UPPAAL is needed in order to benefit from ANIMO.
Nevertheless, the \tas\ model and the model checking process in UPPAAL can be accessed when desired by the user.

The dynamic behaviour of a model can be interactively explored by
moving a time slider underneath the graph to highlight time points in a simulation. In the network view,
each node will be coloured according to its activity level at the selected time point.
Experimental data can be compared to the model by importing and superposing these data
upon an output graph from the model (Figure~\ref{fig:small-model} B,D,F). The ANIMO user workflow and the
features described above are illustrated in Suppl. Video 1.

@@

\begin{figure}[htb]
\begin{minipage}{\textwidth}
  \includegraphics[width=\textwidth]{images/animo_simulation_workflow}
\caption{The passages intercurring between the press of the \emph{Analyse network} button
and the generation of an activity level graph.
A simulation produces, for each selected node, a series of pairs {\sf ($t$, $a$)},
with $t$ a time instant along the simulation and $a$ an activity level. These data are
then parsed and translated into a graph.}\label{fig:animo-sim-workflow}
\end{minipage}
\end{figure}

@@

@@

\begin{figure}[htpb]
\begin{minipage}{\textwidth}
\centering
  \includegraphics[width=.7\textwidth]{images/large_network_tnfa2}
\caption{The model for the TNF$\alpha$ pathway in isolation. Node colours represent the activity level of the
corresponding modelled reactants at time $t = 10$ minutes after a stimulation of 100 ng/ml TNF$\alpha$.}\label{fig:large-model-tnf}
\end{minipage}
\end{figure}

>>>>

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}
\pgfdeclareimage[width=0.95\textwidth]{100TNF0EGF}{images/TNFa-EGF/100TNFa-0EGF/100TNF_0EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{0TNF100EGF}{images/TNFa-EGF/0TNFa-100EGF/0TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{100TNF100EGF}{images/TNFa-EGF/100TNFa-100EGF/100TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{CircadianComparison}{images/Circadian/ANIMOvsODE-crop}
\pgfdeclareimage[width=0.9\textwidth]{ANIMOvsODE-smaller-comparison}{images/Circadian/ANIMOvsODE_small_comparison-crop}

@@

The output of the ANIMO model (see Suppl. Fig.~\ref{fig:drosophila-model}) matches the original ODE model.
In particular, the oscillations in both models show similar periods and phases (see Fig.~\ref{fig:grafici-drosophila-smaller} for
an example, and Suppl. Fig~\ref{suppl-fig:grafici-drosophila} for the complete comparison).
% Due to the compositional nature of \tas\, ANIMO allows for intuitive \emph{in silico} knock-out experiments,
% by right-clicking a node in the model and disabling it. Such experiments have been done
% before~\cite{drosophila-ode-model} and give similar results in our model.
Details of the comparison
between the ANIMO model and the original ODE model are given in Supplementary Section~\ref{suppl-sec:animo-drosophila}.
% \\
% {\Large [[Insert description of K.O. experiments with Drosophila model]]}

\begin{figure*}[!htpb]
\begin{center}
\pgfuseimage{ANIMOvsODE-smaller-comparison}
\end{center}
\caption{\label{fig:grafici-drosophila-smaller}\csentence{Comparing the result of the ANIMO model of \emph{Drosophila Melanogaster} circadian clock
with the model of~\cite{drosophila-ode-model}.} 24 hours simulations of the two models were compared
against each other, synchronizing their start point as much as possible. The blue line is the ANIMO model
({\sf \_{}ANIMO} series), while the red line represents the data computed from the original ODE model
({\sf \_{}ODE} series). The vaules of all series were rescaled on a [0, 100] interval to facilitate comparison.}
\end{figure*}

@@

\def\largeModelScale{0.18}%0.15}%0.155}%0.27}
\def\legendScalaColori{0.21}%0.18}%0.21}
\def\legendScalaForme{0.21}%0.18}%0.21}
\def\scalaGrafici{0.0709}%0.13}
\begin{figure*}[!htpb]
\centering
%  \subfloat[\label{fig:large-model-no-hypotheses}]{\includegraphics[scale=\largeModelScale]{images/large_network_merged_no_hypotheses}}\\
% %\subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme}}\\ \subfloat{\includegraphics[scale=\legendScalaForme]{images/legenda_colori}}\\
% \subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme_e_colori}}\\
% \addtocounter{subfigure}{-1}
% \subfloat[\label{fig:large-model-graph1}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_TNF0_ho_hyp_IL-1a3}}
% \subfloat[\label{fig:large-model-graph3}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_no_hyp_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_merged_no_hypotheses_completa}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-no-hypotheses}}
The model for the merged TNF$\alpha$ and EGF pathways. Node colours represent the
activity level of the corresponding modelled reactants at time $t = 15$ minutes after
a stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph1}}~
Modelled production of IL-1$\alpha$ after stimulation with 100 ng/ml TGF$\alpha$ (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph3}}~
Modelled activation of JNK1 and MK2 after stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-all}
\end{figure*}

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_completa}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

There are other nodes for which the experimental data deviates from the model in one or more of the experimental conditions.
A comparison between model and experimental data can be found in Supplementary Figures~\ref{fig:differences1}, \ref{fig:differences2} and~\ref{fig:differences3}.
Comparing these results with the ones from~\cite{pathway-fuzzy} shows a better fit of the ANIMO model, which is also intrinsically
more precise, being more mechanistic in nature (see Fig.~\ref{fig:animo-spectrum}).
A complete deciphering of the signalling events
in this biological system is outside the scope of this paper. Instead, we illustrated how interactive modelling of the dynamic behaviour
of a signal transduction network can be used to extend previous pathway topologies and can lead to the generation of novel hypotheses.

@@

Timed Automata have been shown to be a powerful formalism to model biological processes
~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together
these locations represent the current state of the biological system. Each timed automaton can have one or more local clocks
associated to it, allowing temporal control of transitions between locations. These transitions are used to
represent interactions between molecules. Fast interactions take less time than slow interactions
to perform an activation or inhibition step. We have previously described in detail how approximated
scenarios~\cite{animo-ieee} (see also Suppl. Sect.~\ref{subsec:abstractions}) can be used to calculate the timing of molecular
interactions to give a description of network dynamics. Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas.
This model describes the activation of ERK by MEK.

@@

\begin{figure*}[!htbp]
\centering
%  \\
% \subfloat{\includegraphics[scale=\legendGraphScale]{images/small-model-1g_legenda}}\addtocounter{subfigure}{-1}\subfloat[\label{fig:small-model-first}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model1f}}
% %\subfloat[\label{fig:small-model-second}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model2f}}
% \subfloat[\label{fig:small-model-third}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model3f}}
% \subfloat[\label{fig:small-model-fourth}]{\includegraphics[scale=\modelGraphScale]{images/00-paper-model4g}}
%    \\
% \subfloat[\label{fig:small-model-first-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph1m_riga}}
% %\subfloat[\label{fig:small-model-second-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph2m_riga}}
% \subfloat[\label{fig:small-model-third-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph3n_riga}}
% \subfloat[\label{fig:small-model-fourth-graph}]{\includegraphics[scale=\halfGraphScale]{images/00-paper-graph4o_riga}}
  \includegraphics[width=0.9\textwidth]{images/small_model_completa}
  \caption{
%{\bf (\protect\subref*{subfig:mek-erk} - \protect\subref*{subfig:mek})}
%\\
%{\bf (\protect\subref*{fig:small-model-first} - \protect\subref*{fig:small-model-fourth-graph})}
\csentence{Construction of an ANIMO model of signal transduction
events in human colon carcinoma cells upon stimulation with 100 ng/ml TNF$\alpha$.}
Graphs below show the dynamic behaviour of the corresponding models above, comparing it to the measured
activity values from~\cite{pathway-compendium} (error bars represent the standard deviation).
On the vertical axis, ``100'' represents the maximum protein activity in the complete experiment.
A red vertical line in each graph highlights an arbitrary time point in the time course:
nodes in the corresponding model are coloured according to their activity at that time point.
{\bf(A,~D)} %{\bf (\protect\subref*{fig:small-model-first}, \protect\subref*{fig:small-model-first-graph})}~
Basic model showing direct activation of JNK1 and MK2 by TNF$\alpha$.
No peak dynamics are observed because no inactivating processes are present.
%{\bf (\protect\subref*{fig:small-model-second}, \protect\subref*{fig:small-model-second-graph})}~The model with added phosphatases.
{\bf(B,~E)} %{\bf (\protect\subref*{fig:small-model-third}, \protect\subref*{fig:small-model-third-graph})}~
The model after addition of inactivating phosphatases and a negative feedback loop that down-regulates TNFR. Note that adding TNFR internalization or phosphatases alone would not be enough to reproduce activity peaks.
{\bf(C,~F)} %{\bf (\protect\subref*{fig:small-model-fourth}, \protect\subref*{fig:small-model-fourth-graph})}~
The model after addition of IKK, IL1-secretion (abstracting
the autocrine IL-1 signalling described in~\cite{pathway-autocrine}), Casp8 and Casp3, showing the late response to TNF$\alpha$ signalling.
As the data set did not contain values for cleaved caspase-3, but only for its non-cleaved precursor pro-caspase-3,
we computed the {\sf Casp3\_{}data} series as $100\% - [\mbox{\sf pro-Casp3}]$.\label{fig:small-model}}
\end{figure*}
%\end{floatingfigure}

@@

\begin{center}
\begin{longtable}{lll}
\caption{Explanation of the abbreviated names referring to molecular species
in the main text.}\label{suppl-tab:names} \\
%\begin{tabular}{lll}
\hline
\noalign{\vskip 2mm} {\bfseries Abbreviation} & {\bfseries Full name} & {\bfseries UniProt ID}\\[2mm]
\hline
\endfirsthead

\multicolumn{3}{c}%
{\scriptsize{\bfseries \tablename\ \thetable{}} \--\ continued from previous page} \\
\hline
\noalign{\vskip 2mm} {\bfseries Abbreviation} & {\bfseries Full name} & {\bfseries UniProt ID}\\[2mm]
\hline
\endhead

@@

Detailed biochemical reaction mechanisms are abstracted to \emph{interactions}, which can
represent either activations ($\rightarrow$) or inhibitions ($\dashv$\hspace{0.1em}).
This aggregation of elementary reactions into single interaction steps reduces the number of kinetic
parameters involved, while preserving cause-and-effect relationships.
For example, consider a reaction in which enzyme $E$ phosphorylates and activates substrate $S$,
transferring a phosphate group from a molecule of ATP to a molecule of $S$. Biochemically, this reaction
can be represented as
$$
\mbox{\it E} + \mbox{\it S} + \mbox{\it ATP} \rightleftarrows \mbox{\it ES} + \mbox{\it ATP} \rightarrow \mbox{\it ES}^{\mbox{\scriptsize \it P}} + \mbox{\it ADP} \rightleftarrows \mbox{\it E} + \mbox{\it S}^{\mbox{\scriptsize\it P}} + \mbox{\it ADP},
$$
with conservation condition $\mbox{\it S} + \mbox{\it S}^{\mbox{\scriptsize\it P}} = \mbox{constant}$ and $\mbox{\it ATP} + \mbox{\it ADP} = \mbox{constant}$.\\
Under the assumption of ATP constantly being replenished by the cell, this reaction is abstracted in ANIMO to the corresponding interaction
$$
\mbox{\it E} \rightarrow \mbox{\it S}.
$$
Each occurrence of the interaction $E \rightarrow S$ will increase the activity level of $S$ by one discrete step.
Since the activity level is defined as the active fraction of a molecular species, an increase in the active fraction
implies a decrease in the inactive fraction. Hence, the original conservation condition is automatically
satisfied.
The interaction rate, $R$, depends on the activity levels of the reactants involved and on a single kinetic
parameter $k$ that is set by the user.
The three available interaction scenarios can be interpreted as abstracted kinetic rate laws:
\begin{enumerate}
  \item $R = k \times [E]$: the interaction rate depends only on the activity level of the upstream node.
  \item $R = k \times [E] \times [1 - S]$ (activations) or $R = k \times [E] \times [S]$ (inhibitions): the rate
  depends on the activity levels of both the upstream and downstream participants. Activations depend on the
  presence of inactive substrate, \emph{[1 - S]}, whereas inhibitions depend on the level of active substrate,
  \emph{[S]}.
  \item $R = k \times [E_1] \times [E_2]$: this scenario can be used when the activation or inhibition
  of a downstream node depends on the simultaneous activity of two upstream nodes. This scenario is comparable to an
  \emph{AND-gate} in Boolean logic.
\end{enumerate}
As we have shown in the Results Section and in~\cite{animo-ieee,animo-gene,animo-syncop}, the abstraction described here preserves ample
expressivity to capture the dynamic behaviour of a biological network.

\subsection{ANIMO}
The modelling approach described in Section~\ref{subsec:abstractions} and in the Methods Section
is implemented in the
software tool ANIMO (Analysis of Networks with Interactive MOdelling, \cite{animo-ieee}) as
a plug-in to the network visualization tool Cytoscape~\cite{cytoscape}. The visual interface of Cytoscape
makes the construction, expansion and rewiring of a network topology a fast and user-friendly process.

@@

ANIMO automatically translates the user input to a \tas\ model, which is then simulated with the model
checking tool UPPAAL~\cite{uppaal}. The results are subsequently parsed and translated to a graph that shows
the dynamic behaviour of nodes in the network.
A schematic overview of this process is given in Figure~\ref{fig:animo-sim-workflow}.
No training or prior knowledge on the use of \tas\ or UPPAAL is needed in order to benefit from ANIMO.
Nevertheless, the \tas\ model and the model checking process in UPPAAL can be accessed when desired by the user.

\begin{figure}[htb]
\begin{center}
\begin{minipage}{\textwidth}
\begin{center}
  \includegraphics[width=0.9\textwidth]{images/animo_simulation_workflow}
\caption{\csentence{The passages intercurring between the press of the \emph{Analyse network} button
and the generation of an activity level graph.}
The model represented in the Cytoscape visual interface is automatically translated by ANIMO
into a network of Timed Automata, which is then simulated using UPPAAL.
A simulation produces, for each selected node, a series of pairs {\sf ($t$, $a$)},
with $t$ a time instant along the simulation and $a$ an activity level. These data are
then parsed and represented as a graph, which is shown in ANIMO's user interface.}\label{fig:animo-sim-workflow}
\end{center}
\end{minipage}
\end{center}
\end{figure}

The dynamic behaviour of a model can be interactively explored by
moving a time slider underneath the graph to highlight time points in a simulation. In the network view,
each node will be coloured according to its activity level at the selected time point.
Experimental data can be compared to the model by importing and superposing these data
upon an output graph from the model (all the graphs in this paper were made directly in ANIMO).
The ANIMO user workflow and the
features described above are illustrated in a video on ANIMO's web page \url{http://fmt.cs.utwente.nl/tools/animo}.

@@

@@

\subsection{ANIMO model of the Drosophila circadian clock}
The model built in ANIMO to represent the circadian clock of \emph{Drosophila Melanogaster}
can be seen in Figure~\ref{fig:drosophila-model}. Its topology is based on the one presented
in~\cite{drosophila-ode-model}, while the parameters were set by hand in order to obtain
an oscillatory behavior.

@@

\begin{figure}[htpb]
\begin{minipage}{\textwidth}
\centering
  \includegraphics[width=.7\textwidth]{images/large_network_tnfa2}
\caption{\csentence{The model for the TNF$\alpha$ pathway in isolation.} Node colours represent the activity level of the
corresponding modelled reactants at time $t = 10$ minutes after a stimulation of 100 ng/ml TNF$\alpha$.}\label{fig:large-model-tnf}
\end{minipage}
\end{figure}
